Hydra Battles and Cie

### 2021-03-18 doc/chapter-primrec.tex
consistent use of capitalization in headers
add variants of Ackermann function + a few typos

\chapter{Primitive Recursive Functions}

@@

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

The following proof decomposes the \texttt{double} function as the composition of 
multiplication,with the pair composed of the identity and the constant function wihich returns $2$.
\emph{Note that the lemma \texttt{const1\_NIsPR} considers this function as an unary function (unlike \texttt{const0\_NIsPR} }.

@@

\index{primrec}{Exercises}
\begin{exercise}
Consider the following functions:

@@

Prove they are primitive recursive.

\textbf{Hint:} You may have to look again at the lemmas of the library
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec} if you meet some difficulty.

\emph{You may start this exercise with the file
\url{../exercises/primrec/MorePRExamples.v}.}
\end{exercise}

@@

>>>>

\chapter{Primitive recursive functions}

@@

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

The following proof decomposes the \texttt{double} function as the composition of 
multiplication with the identity and the constant function wihich returns $2$.
\emph{Note that the lemma \texttt{const1\_NIsPR} considers this function as an unary function (unlike \texttt{const0\_NIsPR})}.

@@

\index{primrec}{Exercises}
\begin{exercise}
Prove that the following functions are primitive recursive.

@@

\textbf{Hint:} You may have to look again at the lemmas of the library
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec} if you meet some difficulty.
You may start this exercise with the file
\url{../exercises/primrec/MorePRExamples.v}.
\end{exercise}

@@

\index{hydras}{Exercises}

\begin{exercise}
The file \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack} presents two other definitions of the Ackermann functions based on the lexicographic ordering on $\mathbb{N}\times\mathbb{N}$.
Prove that all the four functions are extensionnally equal.
\end{exercise}

### 2021-03-18 doc/part-hydras.tex
add variants of Ackermann function + a few typos

Simply put, this document tries to combines the scientific interest of two articles~\cite{KP82, KS81} and a book~\cite{schutte} with the playful activity of truly proving theorems.
We hope  that such a work, besides exploring a nice piece of discrete maths, 
will show how \coq{} and its standard library are well fitted to help us to understand some non-trivial mathematical developments, and also to experiment the constructive parts of  the proof through functional programming.

 We also hope to provide a little clarification on infinity (both potential and actual) through the notions of function, computation, limit,
 types and proofs.

@@

\subsection*{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

@@

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}). 
\emph{For a less formal description of hydras, please see 
\url{https://www.smbc-comics.com/comic/hydra}.}

@@

\item Please keep in mind that the last  configuration of your interactively built battle is known only at the end of the battle. Thus, you will have to create and solve subgoals with existential variables. For that purpose, the tactic \texttt{eexists}, applied to the 
goal \Verb@{h':Hydra | h -*-> h'}@ generates the subgoal \Verb|h -*-> ?h'|.
\item You may use Gérard Huet's \emph{zipper} data structure~\cite{zipper} for writing tactics associated with Hercule's  interactive search of a head to chop off.
\end{itemize}

@@

Class Battle :=  {battle_r : dep_round_t;
                  battle_inclusion : forall i h h',
                      battle_r i h h' -> round h h'}.

@@

\begin{Coqsrc}
Program Instance free : Battle :=
  (Build_Battle ( fun _  h h' => round h h') _).
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive battle (B:Battle) : nat -> Hydra -> nat -> Hydra -> Prop :=
| battle_1 : forall i h  h', 
      battle_r   B i  h h' -> battle B i h (S i) h'
| battle_n : forall i h  j h' h'',  
      battle_r  B i h h''  ->
      battle B (S i) h'' j h'  ->
      battle B i h j h'.
\end{Coqsrc}

@@

\index{hydras}{Library Hydra!Type classes!Hvariant}
\begin{Coqsrc}
Class Hvariant {A:Type}{Lt:relation A}(Wf: well_founded Lt)(B : Battle)
  (m: Hydra -> A):   Prop :=
  {variant_decr :forall i h h',
      h <> head ->
      battle_r  B i  h h' -> Lt (m h') (m h)}.
\end{Coqsrc}

>>>>

Simply put, this document tries to combine the scientific interest of two articles~\cite{KP82, KS81} and a book~\cite{schutte} with the playful activity of truly proving theorems.
We hope  that such a work, besides exploring a nice piece of discrete maths, 
will show how \coq{} and its standard library are well fitted to help us to understand some non-trivial mathematical developments, and also to experiment the constructive parts of  the proof through functional programming.

 We also hope to provide a little clarification on infinity (both potential and actual) through the notions of function, computation, limit,
 type and proof.

@@

\subsection*{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris show  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers up to $\epsilon_0$.

@@

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Contrary to the computer science tradition, we display the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}). 
\emph{For a less formal description of hydras, please see 
\url{https://www.smbc-comics.com/comic/hydra}.}

@@

\item Please keep in mind that the last  configuration of your interactively built battle is known only at the end of the battle. Thus, you will have to create and solve subgoals with existential variables. For that purpose, the tactic \texttt{eexists}, applied to the 
goal \Verb@{h':Hydra | h -*-> h'}@ generates the subgoal \Verb|h -*-> ?h'|.
\item You may use Gérard Huet's \emph{zipper} data structure~\cite{zipper} for writing tactics associated with Hercule's  interactive search for a head to chop off.
\end{itemize}

@@

Class Battle :=  
    { battle_rel : dep_round_t;
      battle_inclusion : forall i h h', battle_rel i h h' -> round h h'}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Program Instance free : Battle := Build_Battle (fun _  h h' => round h h') _.
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive battle (B:Battle) : nat -> Hydra -> nat -> Hydra -> Prop :=
| battle_1 : forall i h  h', 
      battle_rel   B i  h h' -> battle B i h (S i) h'
| battle_n : forall i h  j h' h'',  
      battle_rel  B i h h''  ->
      battle B (S i) h'' j h'  ->
      battle B i h j h'.
\end{Coqsrc}

@@

\index{hydras}{Library Hydra!Type classes!Hvariant}
\begin{Coqsrc}
Class Hvariant {A:Type}{Lt:relation A}(Wf: well_founded Lt)(B : Battle)
  (m: Hydra -> A):   Prop :=
  {variant_decr: forall i h h',
      h <> head ->
      battle_rel  B i  h h' -> Lt (m h') (m h)}.
\end{Coqsrc}
